#### load libraries

```{r}
library(tidyverse)
library(tximport)
library(DESeq2)
library(readxl)
library(PCAtools)
source("/project/imoskowitz/shared/software/helper_functions/carlosFunc_3.R")
source("/project/imoskowitz/shared/software/helper_functions/rowr.R")

```

#### create directories
```{r}
dir.create("pdfs", mode = "750")
dir.create("data", mode = "750")
dir.create("code", mode = "750")
```

#### import data
```{r}
txi_RC7 <- salmon_import("/project/imoskowitz/shared/sequencing.processed/20241025_LH00315_0309_A22NVHHLT3-XRS40-IM-RS21/star_salmon")
txi_RC1 <- salmon_import("/project/imoskowitz/shared/sequencing.processed/RC1/star_salmon")
```

#### import metadata
```{r}

metadata_RC7 <- load_metadata("/project/imoskowitz/shared/metasheets/sample_sheets/RC07_mESCCPs_Sample_metasheet_v17.xlsx")
metadata_RC1 <- load_metadata("/project/imoskowitz/shared/metasheets/sample_sheets/Sample_RC1_bulkRNAseq_Gli1FTA.xlsx")

metadata_RC1_noBMP <- subset(metadata_RC1, grepl("NoBMP4", metadata_RC1$Sample.name))

txi_RC1_noBMP <- lapply(txi_RC1[1:3], function(x) x[ , colnames(x) %in% rownames(metadata_RC1_noBMP)])

metadata_RC1_noBMP <- metadata_RC1_noBMP[match(colnames(txi_RC1_noBMP[[2]]), rownames(metadata_RC1_noBMP)), ]
metadata_RC7 <- metadata_RC7[match(colnames(txi_RC7[[2]]), rownames(metadata_RC7)), ]

txi_RC1_noBMP$countsFromAbundance <- "no"
```

#### run DESeq2 and pulll raw counts
```{r}
metadata_RC7$batch <- str_extract(string = metadata_RC7$Sample.name, pattern = "DC02[0-9]")

dds_RC1_noBMP <- DESeqDataSetFromTximport(txi_RC1_noBMP, metadata_RC1_noBMP, design = ~ time_condition)
dds_RC7 <- DESeqDataSetFromTximport(txi_RC7, metadata_RC7, design = ~ batch+time_condition)

RC7_counts <- counts(dds_RC7, normalized = FALSE)
RC1_counts <- counts(dds_RC1_noBMP, normalized = FALSE)

```

#### get average raw counts per timepoint/condition for RC1 and also get collaspsereps
```{r}
RC1_collapsed <- collapseReplicates(dds_RC1_noBMP, dds_RC1_noBMP$time_condition, renameCols = TRUE) %>%
counts(normalized = FALSE)

RC1_counts_avg <- reshape2::melt(RC1_counts) %>%
  merge(metadata_RC1_noBMP[7], by.x = "Var2", by.y = 0) %>%#
  group_by(Var1, time_condition) %>%
  summarise(mean = mean(value)) %>%
  reshape2::dcast(Var1 ~ time_condition, value.var = "mean") %>%
  column_to_rownames("Var1") %>% round()

colnames(RC1_counts_avg) <- paste0(colnames(RC1_counts_avg), "_RC1")
colnames(RC1_collapsed) <- paste0(colnames(RC1_collapsed), "_RC1")

RC1_metadata_combined_reps <- metadata_RC1_noBMP %>%
remove_rownames() %>%
select(Age.stage, treatment,time_condition) %>%
unique()

rownames(RC1_metadata_combined_reps) <- paste0(RC1_metadata_combined_reps$time_condition, "_RC1")

RC1_counts_avg <- select(RC1_counts_avg, all_of(rownames(RC1_metadata_combined_reps)))

```

#### check and correct for batch effects in RC7
```{r}

### try DESeq2 batch correction
dds_RC7 <- DESeqDataSetFromTximport(txi_RC7, metadata_RC7,
design = ~batch+time_condition) %>%
DESeq(.)

RC7_res <-results(dds_RC7 , contrast = c("time_condition", "D6_dox", "D6_ctrl"), alpha = 0.05) %>%
    lfcShrink(dds = dds_RC7,
              res = .,
              contrast = c("time_condition", "D6_dox", "D6_ctrl"),
              type = 'normal') %>%
summary()

#plot GLI1 boxplot 
vst <- varianceStabilizingTransformation(dds_RC7, blind = FALSE)
pca <- PCAtools::pca(assay(vst), metadata = metadata_RC7, removeVar = 0.9)
PCAtools::biplot(pca, colby = "time_condition",
                legendPosition = "right",
                shape = "batch",
                pointSize = 3,
                labSize = 2)

ggsave(filename = "pca_biplot_time_condition_RC7_DESeq2_batch_correction.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()

assay(vst) %>% reshape2::melt() %>%
subset(Var1 == "Gli1") %>%
merge(.,metadata_RC7, by.x = "Var2", by.y= 0) %>%
ggplot(aes(y = value, x = Age.stage, color = time_condition)) +
geom_boxplot(outlier.shape = NA) +
ggrastr::geom_beeswarm_rast(size = 1) +
ggtitle("Gli1") +
theme_minimal(base_size = 15)

ggsave(filename = "Gli1_boxplot_RC7_DESeq2_batch_correction.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()

### try combatseq


dds_RC7 <-counts(dds_RC7) %>% as.matrix() %>%
sva::ComBat_seq(., batch = metadata_RC7$batch, group = metadata_RC7$time_condition) %>%
DESeqDataSetFromMatrix(., metadata_RC7, design = ~batch+time_condition) %>%
DESeq(.)

results(dds_RC7 , contrast = c("time_condition", "D6_dox", "D6_ctrl"), alpha = 0.05) %>%
    lfcShrink(dds = dds_RC7,
              res = .,
              contrast = c("time_condition", "D6_dox", "D6_ctrl"),
              type = 'normal') %>%
summary()

vst_combat <- varianceStabilizingTransformation(dds_RC7, blind = FALSE)
pca_combat <- PCAtools::pca(assay(vst_combat), metadata = metadata_RC7, removeVar = 0.9)
PCAtools::biplot(pca_combat, colby = "time_condition",
                legendPosition = "right",
                shape = "batch",
                pointSize = 3,
                labSize = 2)
ggsave(filename = "pca_biplot_time_condition_RC7_combatseq.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()

assay(vst_combat) %>% reshape2::melt() %>%
subset(Var1 == "Gli1") %>%
merge(.,metadata_RC7, by.x = "Var2", by.y= 0) %>%
ggplot(aes(y = value, x = Age.stage, color = time_condition)) +
geom_boxplot(outlier.shape = NA) +
ggrastr::geom_beeswarm_rast(size = 1) +
ggtitle("Gli1") +
theme_minimal(base_size = 15)

ggsave(filename = "Gli1_boxplot_RC7_combatseq.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()


```




#### merge RC1 and RC7 counts while correcting for batch effects between the two with DESeq2 (soft correction)
```{r}
RC7_metadata_s <- select(metadata_RC7, Age.stage, treatment,batch,time_condition) 
RC7_metadata_s$batch_exp <- "0"
RC7_metadata_s$Sample_name <- paste0( RC7_metadata_s$time_condition, "_", rownames(RC7_metadata_s)) %>%
gsub(pattern = "RC7_R", replacement = "R")

idx <- match(colnames(RC7_counts), rownames(RC7_metadata_s))
colnames(RC7_counts)[idx] <- RC7_metadata_s$Sample_name[idx]
rownames(RC7_metadata_s) <- RC7_metadata_s$Sample_name


RC1_metadata_combined_reps$batch_exp <- "1"
RC1_metadata_combined_reps$batch <- "n"
RC1_metadata_combined_reps$Sample_name <- rownames(RC1_metadata_combined_reps)
RC1_RC7_metadata <- rbind(RC1_metadata_combined_reps, RC7_metadata_s)

RC1_RC7_metadata$time_condition <- gsub(pattern = "oe", 
replacement = "dox", x = RC1_RC7_metadata$time_condition)


RC1_RC7_counts <- merge(RC1_counts_avg, RC7_counts, by =0) %>%
  column_to_rownames("Row.names") %>%
  select(all_of(rownames(RC1_RC7_metadata)))

dds_RC1_RC7 <- DESeqDataSetFromMatrix(RC1_RC7_counts, 
RC1_RC7_metadata, design = ~ time_condition+batch) %>%
DESeq(.)
vst_RC1_RC7 <- varianceStabilizingTransformation(dds_RC1_RC7, blind = FALSE)
### check batch with PCA, heatmap, and boxplots
#plot GLI1 boxplot 
vst_assay <- assay(vst_RC1_RC7)

vst_assay %>% reshape2::melt() %>%
subset(Var1 == "Gli1") %>%
merge(., RC1_RC7_metadata, by.x = "Var2", by.y= 0) %>%
mutate(Age.stage = factor(Age.stage, levels = c("D5","D5.6h", "D6", "D7","D8","D12"))) %>%
ggplot(aes(y = value, x = Age.stage, color = time_condition)) +
geom_boxplot(outlier.shape = NA) +
ggrastr::geom_beeswarm_rast(size = 1) +
ggtitle("Gli1") +
theme_minimal(base_size = 15)

ggsave(filename = "Gli1_boxplot_RC1_RC7_DESeq2_batch_correction.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()

pca <- PCAtools::pca(vst_assay, metadata = RC1_RC7_metadata, removeVar = 0.9)

PCAtools::biplot(pca, colby = "time_condition",
                legendPosition = "right",
                shape = "batch_exp",
                pointSize = 3,
                labSize = 2)

ggsave(filename = "pca_biplot_time_condition_RC1_RC7.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()
### check only day 8 in PCA
vst_assay_day8 <- vst_assay[ , RC1_RC7_metadata$Age.stage == "D8" ]

pca_day8 <- PCAtools::pca(vst_assay_day8, metadata = RC1_RC7_metadata[RC1_RC7_metadata$Age.stage == "D8", ], removeVar = 0.9)
PCAtools::biplot(pca_day8, colby = "time_condition",
                legendPosition = "right",
                shape = "batch_exp",
                pointSize = 3,
                labSize = 2)

ggsave(filename = "pca_biplot_time_condition_RC1_RC7_D8.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()


pdf(file = "pdfs/heatmap_sample_cor_RC1_RC7.pdf", height = 7, width = 14)

vst_assay %>%
  cor() %>%
  pheatmap::pheatmap(scale = "none",
    display_numbers = T,
    annotation_row = RC1_RC7_metadata[c("time_condition", "batch_exp")],
    annotation_col = RC1_RC7_metadata[c("time_condition", "batch_exp")])

dev.off()

saveRDS(dds_RC1_RC7, file = "data/dds_RC1_RC7_DESeq2_batch_correction.rds")

```


#### try the batch correction with combatseq
```{r}

RC1_RC7_counts_combat <- sva::ComBat_seq( as.matrix(RC1_RC7_counts),
batch = RC1_RC7_metadata$batch_exp,
group = RC1_RC7_metadata$time_condition)

dds_RC1_RC7_combat <- DESeqDataSetFromMatrix(RC1_RC7_counts_combat, 
RC1_RC7_metadata, design = ~ time_condition)

dds_RC1_RC7_combat <- DESeq(dds_RC1_RC7)
vst_RC1_RC7_combat <- varianceStabilizingTransformation(dds_RC1_RC7_combat, blind = FALSE)


vst_assay_combat <- assay(vst_RC1_RC7_combat)
pca_combat <- PCAtools::pca(vst_assay_combat, metadata = RC1_RC7_metadata, removeVar = 0.9)

PCAtools::biplot(pca_combat, colby = "time_condition",
                legendPosition = "right",
                shape = "batch_exp",
                pointSize = 3,
                labSize = 2)

ggsave(filename = "pca_biplot_time_condition_RC1_RC7_combat.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()
### check only day 8 in PCA
vst_assay_day8_combat <- vst_assay_combat[ , RC1_RC7_metadata$Age.stage == "D8" ]

pca_day8_combat <- PCAtools::pca(vst_assay_day8_combat, metadata = RC1_RC7_metadata[RC1_RC7_metadata$Age.stage == "D8", ], removeVar = 0.9)
PCAtools::biplot(pca_day8_combat, colby = "time_condition",
                legendPosition = "right",
                shape = "batch_exp",
                pointSize = 3,
                labSize = 2)

ggsave(filename = "pca_biplot_time_condition_RC1_RC7_D8_combat.pdf",
       plot = last_plot(),
       device = "pdf",
       width = 8, height = 5,
       path = "pdfs")
dev.off()


pdf(file = "pdfs/heatmap_sample_cor_RC1_RC7_combat.pdf", height = 7, width = 14)

vst_assay_combat %>%
  cor() %>%
  pheatmap::pheatmap(scale = "none",
    display_numbers = T,
    annotation_row = RC1_RC7_metadata[c("time_condition", "batch_exp")],
    annotation_col = RC1_RC7_metadata[c("time_condition", "batch_exp")])

dev.off()


saveRDS(dds_RC1_RC7_combat, file = "data/dds_RC1_RC7_combat.rds")
```


#### perform various contrats and save results
```{r}
my_contrasts <- list(
D5.6h_dox_vs_D5.6h_ctrl = c("time_condition", "D5.6h_dox", "D5.6h_ctrl"),
D6_dox_vs_D6_ctrl = c("time_condition", "D6_dox", "D6_ctrl"),
D7_dox_vs_D7_ctrl = c("time_condition", "D7_dox", "D7_ctrl"),
D8_dox_vs_D8_ctrl = c("time_condition", "D8_dox", "D8_ctrl"),
D12_dox_vs_D12_ctrl = c("time_condition", "D12_dox", "D12_ctrl"))

deseq_results <- list()

for (contrast in names(my_contrasts)) {
    deseq_results[[contrast]] <- get_deseq_results(dds_RC1_RC7,
    contrast = my_contrasts[[contrast]][1], 
    treatment = my_contrasts[[contrast]][2],
    control = my_contrasts[[contrast]][3])

write.csv(deseq_results[[contrast]],
file = paste0("data/Table_deseq2_results_", contrast, ".csv"))

}


# Plot volcanos and save results
for (i in 1:length(deseq_results)) {
  res <- deseq_results[[i]]
  gg_volcano3(res,
              abs_lfc = 0,
              p.cutoff = 0.05,
              up.color = "darkseagreen",
              down.color = "plum4") +
    gg_geneLabel4(res,
                  gene_col = "gene_id",
                  colorText = "black",
                  plotGenes = c("Foxf1",'Gli1', 'Hhip', 'Ptch1')) +
    theme_bw(base_size = 16) +
    annotate(
      geom = "text",
      x = c(-Inf, Inf),
      y = c(Inf, Inf),
      hjust = c(-0.5, 1.5),
      vjust = c(2, 2),
      label = c(nrow(subset(res, log2FoldChange < 0 & padj < 0.05)),
                nrow(subset(res, log2FoldChange > 0 & padj < 0.05))),
      size = 6
    ) +
    ggtitle(names(deseq_results)[i])
  
  ggsave(filename = paste0("volcano_", names(deseq_results)[i], ".pdf"),
         plot = last_plot(),
         device = "pdf",
         width = 5, height = 5,
         path = "pdfs")
}

# Run clusterProfiler
go_df_up <- list()

for (i in 1:length(deseq_results)) {
  tryCatch({
    res <- deseq_results[[i]]
    
    x <- runGO(test.vector = subset(res, padj < 0.05 & log2FoldChange > 0)$gene_id,
               univr_vector = res$gene_id,
               ont = "BP",
               organism = 'mouse',
               simplify = TRUE)
    
    plotGO(obj = x,
           topTerms = 5,
           muhcolor = "darkseagreen",
           strWrap = 30,
           barWidth = 0.5,
           expandY = c(0.1,0))
    
    ggsave(filename = paste0("GO_BP_", names(deseq_results)[i], "_up.pdf"),
           plot = last_plot(),
           device = "pdf",
           width = 7, height = 5,
           path = "pdfs")
    
    go_df_up[[i]] <- x
    names(go_df_up)[i] <- paste0("GO_BP_", names(deseq_results)[i], "_up")
    
  }, error = function(e){cat("ERROR :", conditionMessage(e), "\n")})
}

# Now down
go_df_down <- list()
for (i in 1:length(deseq_results)) {
  tryCatch({
    res <- deseq_results[[i]]
    
    y <- runGO(test.vector = subset(res, padj < 0.05 & log2FoldChange < 0)$gene_id,
               univr_vector = res$gene_id,
               ont = "BP",
               organism = 'mouse',
               simplify = TRUE)
    
    plotGO(obj = y,
           topTerms = 5,
           muhcolor = "plum4",
           strWrap = 30,
           barWidth = 0.5,
           expandY = c(0.1,0))
    
    ggsave(filename = paste0("GO_BP_", names(deseq_results)[i], "_down.pdf"),
           plot = last_plot(),
           device = "pdf",
           width = 7, height = 5,
           path = "pdfs")
    
    go_df_down[[i]] <- y
    names(go_df_down)[i] <- paste0("GO_BP_", names(deseq_results)[i], "_down")
  }, error = function(e){cat("ERROR :", conditionMessage(e), "\n")})
}

go_df_up <- Filter(length, go_df_up)
go_df_down <- Filter(length, go_df_down)

writexl::write_xlsx(
  lapply(go_df_down, data.frame),
  "data/Table_GO_BP_down_padj0.05.xlsx")

writexl::write_xlsx(
  lapply(go_df_up, data.frame),
  "data/Table_GO_BP_up_padj0.05.xlsx")



```