```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(maSigPro))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(PCAtools))

source("/project/imoskowitz/shared/software/helper_functions/carlosFunc_3.R")
source("/project/imoskowitz/shared/software/helper_functions/rowr.R")

RC1_RC7_DESeq2 <- readRDS("./data/dds_RC1_RC7_DESeq2_batch_correction.rds")
RC1_RC7_combat <- readRDS("./data/dds_RC1_RC7_combat.rds")

```

#### prepare data for maSigPro
```{r}

stage_order <- c( "D5", "D5.6h", "D6", "D8")
rep_factor <- colData(RC1_RC7_DESeq2)$time_condition %>%
factor(levels = c("D5_ctrl", "D5.6h_ctrl",
"D6_ctrl", "D8_ctrl")) %>%
as.numeric()
# Apply custom ordering
colData(RC1_RC7_DESeq2) %>% 
data.frame() %>%
subset(., !Age.stage %in% c("mESC", "D4", "D7", "D12")
& !treatment %in% c("dox", "oe")) %>%
dplyr::select(Age.stage, treatment, time_condition) %>%
.[str_order(.$Age.stage, numeric = TRUE),] %>%
mutate(Age.stage = factor(Age.stage, levels = stage_order)) %>%
.[order(.$Age.stage, decreasing = FALSE),] %>%
mutate(Replicate = as.numeric(factor(time_condition,
    levels=c("D5_ctrl", "D5.6h_ctrl","D6_ctrl", "D8_ctrl")) )) %>%
mutate(Time = as.numeric(Age.stage)) %>%
mutate(ctrl = as.numeric(treatment == "ctrl")) %>%
dplyr::select(Time,Replicate, ctrl) -> msp_metadata



```

It looks like DESeq2 does not return batch-correct normalized counts.
They are only used for the dispersion estimation and for foldchange/pvalue calculation.
will do limma voom batch correction followed by combat and compare performance.
#### run maSigPro with limma batch correction
```{r}

xncounts <- counts(RC1_RC7_DESeq2, normalized = TRUE) %>%
data.frame() %>%
dplyr::select(all_of(rownames(msp_metadata))) %>%
+1 %>%
log2()
counts_afterbatchcorrection <- limma::removeBatchEffect(xncounts, RC1_RC7_DESeq2$Batch)


p.vec.obj <- p.vector( counts_afterbatchcorrection, 
    make.design.matrix( msp_metadata, degree = 3), 
    Q = 0.01, MT.adjust = "BH", 
    counts = T) 

t.fit.obj <- T.fit(data=p.vec.obj, 
    step.method = "backward") 


get.siggens.obj <- get.siggenes(t.fit.obj, rsq = 0.6, vars = 'groups') 


```



#### plot elbow plot
```{r, elbow plot}

#Elbow Method for finding the optimal number of clusters
set.seed(123)
# Compute and plot wss for k = 2 to k = 15.
k.max <- 9
data <- counts_afterbatchcorrection
data <- subset(counts_afterbatchcorrection,
rownames(counts_afterbatchcorrection) %in% get.siggens.obj$summary[,1])
wss <- sapply(1:k.max, 
               function(k){kmeans(data, k, nstart=50,iter.max = 15 )$tot.withinss})

plot(1:k.max, wss,
      type="b", pch = 19, frame = FALSE, 
      xlab="Number of clusters K",
      ylab="Total within-clusters sum of squares")
```


Lets go with 9 clusters.
```{r, plot time series and get results, warning=FALSE, message=FALSE}
source("./code/see_genes_edit.R")

pdf("./pdfs/time_series_msp_ctrl_limma_batch_correction.pdf")
see.genes(get.siggens.obj$sig.genes$ctrl, 
    color.mode='gray',cexlab=1,show.fit = F,
    cex=1,k = 9, newX11=F)
dev.off()



see.genes.ob_edit <- see.genes_edit(get.siggens.obj$sig.genes$ctrl, 
    color.mode='gray',cexlab=1,
    cex=1,k = 9, newX11=T)

geneGroups <- see.genes.ob_edit$cut %>% 
data.frame() %>%
rownames_to_column("geneSymbol")
colnames(geneGroups)[2] <- "cluster" 

write.csv(geneGroups, "./data/geneGroups_maSigPro_ctrl_limma_batch_correction.csv")


```


```{r, plot GO for select clusters, warning=FALSE, message=FALSE}

list_of_genes <- list   (
cluster_1ctrl = subset(geneGroups, cluster == 1)$geneSymbol,
cluster_2ctrl = subset(geneGroups, cluster == 2)$geneSymbol,
cluster_3ctrl = subset(geneGroups, cluster == 3)$geneSymbol,
cluster_4ctrl = subset(geneGroups, cluster == 4)$geneSymbol,
cluster_5ctrl = subset(geneGroups, cluster == 5)$geneSymbol,
cluster_6ctrl = subset(geneGroups, cluster == 6)$geneSymbol,
cluster_7ctrl = subset(geneGroups, cluster == 7)$geneSymbol,
cluster_8ctrl = subset(geneGroups, cluster == 8)$geneSymbol,
cluster_9ctrl = subset(geneGroups, cluster == 9)$geneSymbol
)

for (i in 1:length(list_of_genes)) {
    
y <- runGO(test.vector = list_of_genes[[i]],
univr_vector = rownames(counts_afterbatchcorrection),
ont = "BP",
organism = 'mouse',
simplify = TRUE)
    
plotGO(obj = y,
topTerms = 5,
muhcolor = "plum4",
strWrap = 30,
barWidth = 0.5,
expandY = c(0.1,0))
    
ggsave(filename = paste0("GO_BP_maSigPro_ctrl_limma_batch_correction_", names(list_of_genes)[i], ".pdf"),
plot = last_plot(),
device = "pdf",
width = 7, height = 5,
path = "pdfs")

writexl::write_xlsx(
  data.frame(y),
  paste0("data/Table_GO_BP_maSigPro_ctrl_limma_batch_correction_", names(list_of_genes)[i], ".xlsx"))   

}


```


#### run maSigPro with combat batch correction
```{r}


xncounts <- counts(RC1_RC7_combat, normalized = TRUE) %>%
data.frame() %>%
dplyr::select(all_of(rownames(msp_metadata))) %>%
+1 %>%
log2()

p.vec.obj <- p.vector( xncounts, 
    make.design.matrix( msp_metadata, degree = 4), 
    Q = 0.01, MT.adjust = "BH", 
    counts = T) 

t.fit.obj <- T.fit(data=p.vec.obj, 
    step.method = "backward") 

get.siggens.obj <- get.siggenes(t.fit.obj, rsq = 0.6, vars = 'groups') 


pdf("./pdfs/time_series_msp_ctrl_combat_batch_correction.pdf")
see.genes(get.siggens.obj$sig.genes$ctrl, 
    color.mode='gray',cexlab=1,show.fit = F,
    cex=1,k = 9, newX11=F)
dev.off()



see.genes.ob_edit <- see.genes_edit(get.siggens.obj$sig.genes$ctrl, 
    color.mode='gray',cexlab=1,
    cex=1,k = 9, newX11=T)

geneGroups <- see.genes.ob_edit$cut %>% 
data.frame() %>%
rownames_to_column("geneSymbol")
colnames(geneGroups)[2] <- "cluster" 

write.csv(geneGroups, "./data/geneGroups_maSigPro_ctrl_combat_batch_correction.csv")

```


#### plot GO for select clusters; combat batch correction
```{r, plot GO for select clusters, warning=FALSE, message=FALSE}

data <- subset(xncounts,
rownames(xncounts) %in% get.siggens.obj$summary[,1])


list_of_genes <- list   (
cluster_1ctrl = subset(geneGroups, cluster == 1)$geneSymbol,
cluster_2ctrl = subset(geneGroups, cluster == 2)$geneSymbol,
cluster_3ctrl = subset(geneGroups, cluster == 3)$geneSymbol,
cluster_4ctrl = subset(geneGroups, cluster == 4)$geneSymbol,
cluster_5ctrl = subset(geneGroups, cluster == 5)$geneSymbol,
cluster_6ctrl = subset(geneGroups, cluster == 6)$geneSymbol,
cluster_7ctrl = subset(geneGroups, cluster == 7)$geneSymbol,
cluster_8ctrl = subset(geneGroups, cluster == 8)$geneSymbol,
cluster_9ctrl = subset(geneGroups, cluster == 9)$geneSymbol
)

for (i in 1:length(list_of_genes)) {
    
y <- runGO(test.vector = list_of_genes[[i]],
univr_vector = rownames(data),
ont = "BP",
organism = 'mouse',
simplify = TRUE)
    
plotGO(obj = y,
topTerms = 5,
muhcolor = "plum4",
strWrap = 30,
barWidth = 0.5,
expandY = c(0.1,0))
    
ggsave(filename = paste0("GO_BP_maSigPro_ctrl_combat_batch_correction_", names(list_of_genes)[i], ".pdf"),
plot = last_plot(),
device = "pdf",
width = 7, height = 5,
path = "pdfs")

writexl::write_xlsx(
  data.frame(y),
  paste0("data/Table_GO_BP_maSigPro_ctrl_combat_batch_correction_", names(list_of_genes)[i], ".xlsx"))   

}


```
